---
title: "Basics of Data Manipulation in R"
---

### Step 1: Creating the Data Frames

Let's start by creating the three data frames you will use for this assignment. You've already provided a great example of how to create `df_employees` and `df_salaries`. I'll add the third one, `df_performance`.

```{r}
# Create your own data frames
df_employees <- data.frame(
  EmployeeID = 1:5,
  Name = c("Alice", "Bob", "Charlie", "David", "Eva"),
  Department = c("Sales", "IT", "HR", "Marketing", "Finance")
)

df_salaries <- data.frame(
  EmployeeID = c(2, 4, 5, 6, 7),
  Salary = c(50000, 60000, 55000, 65000, 70000)
)

df_performance <- data.frame(
  EmployeeID = c(1, 3, 5, 7),
  Rating = c("Excellent", "Good", "Very Good", "Outstanding")
)
```

### Step 2: Combining Data Frames

#### Using `merge()` in Base R

Base R’s `merge()` function allows you to combine two data frames. You can specify how you want to merge them by using the `all`, `all.x`, or `all.y` parameters.

```{r}
# Inner join (only matching EmployeeID)
merged_inner <- merge(df_employees, df_salaries, by = "EmployeeID")
print(merged_inner)

# Full outer join (keep all rows from both data frames)
merged_full <- merge(df_employees, df_salaries, by = "EmployeeID", all = TRUE)
print(merged_full)

# Left join (all from df_employees, matching from df_salaries)
merged_left <- merge(df_employees, df_salaries, by = "EmployeeID", all.x = TRUE)
print(merged_left)

# Right join (all from df_salaries, matching from df_employees)
merged_right <- merge(df_employees, df_salaries, by = "EmployeeID", all.y = TRUE)
print(merged_right)
```

#### Using `dplyr`'s `full_join()`

If you prefer `dplyr`, the syntax is more intuitive, and it's great for chaining multiple operations.

```{r}
library(dplyr)

# Full join using dplyr
merged_full_dplyr <- full_join(df_employees, df_salaries, by = "EmployeeID")
print(merged_full_dplyr)
```

#### Using `data.table`'s `merge()`

For large datasets, `data.table` is often faster. The syntax is quite similar to base R.

```{r}
library(data.table)

# Convert to data.table
dt_employees <- as.data.table(df_employees)
dt_salaries <- as.data.table(df_salaries)

# Full join using data.table
merged_full_dt <- merge(dt_employees, dt_salaries, by = "EmployeeID", all = TRUE)
print(merged_full_dt)
```

### Step 3: Handling Missing Values

After merging, there might be missing values. To identify them:
  
```{r}
# Find employees missing salary information
missing_salaries <- merged_full[is.na(merged_full$Salary), ]
print(missing_salaries)

# Find salary records without employee details
missing_employees <- merged_full[is.na(merged_full$Name), ]
print(missing_employees)
```

### Step 4: Create a Summary

#### Average Salary per Department

```{r}
# Calculate average salary per department
avg_salary_dept <- merged_full %>%
  group_by(Department) %>%
  summarize(Average_Salary = mean(Salary, na.rm = TRUE))
print(avg_salary_dept) 

```

#### Highest Paid Employee in Each Department

```{r}
# Find highest paid employee in each department
highest_paid <- merged_full %>%
  group_by(Department) %>%
  filter(Salary == max(Salary, na.rm = TRUE))
print(highest_paid) 

```


### Step 5: Challenge - Combine All Three Data Frames

```{r}
# Combine all three data frames
combined_data <- full_join(merged_full_dplyr, df_performance, by = "EmployeeID")
print(combined_data) 

```

### Step 6: Visualization

#### Scatter Plot of Salary vs. Performance Rating

To create the scatter plot, you can use `ggplot2`:
  
```{r}
library(ggplot2)

# Scatter plot of Salary vs. Performance Rating
ggplot(combined_data, aes(x = Rating, y = Salary)) +
  geom_point() +
  labs(title = "Scatter Plot of Salary vs. Performance Rating",
       x = "Performance Rating",
       y = "Salary")
```

#### Bar Plot of Average Salary by Department

```{r}
library(ggplot2)
# Bar plot showing average salary by department
ggplot(avg_salary_dept, aes(x = Department, y = Average_Salary)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Salary by Department",
       x = "Department",
       y = "Average Salary")
```

### Summary

1. **Combining Data Frames**: You practiced using different ways of combining data frames (`merge()`, `full_join()`, etc.). Each method has slight variations, especially in handling missing data.
2. **Handling Missing Values**: You learned to identify rows with missing information.
3. **Summarizing Data**: Calculating averages and filtering rows helps in understanding data trends.
4. **Visualizations**: Visualizations are essential for better data insights. You learned to create scatter plots and bar plots using `ggplot2`.

Feel free to experiment with more functions and methods! Each of these steps will help you become more comfortable with data manipulation and visualization in R. Happy coding!